preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-15 Manuscript/manuscript.Rmd
Discussion polishing

To precisely identify boundary locations, we first constructed a _base-level resolution_ predictor space for a particular chromosome *i*. We then apply our optimally trained RF model on the base-level predictor space to extract the probability vector, $p_{n_{i}}$, denoting each base's probability of being a boundary. Here, $n_{i}$ represents the length of chromosome *i*. A threshold (*t*) is specified on $p_{n_{i}}$, such that base coordinates with $p_{n_{i}} \ge t$ are extracted. Next, density-based spatial clustering of applications with noise (DBSCAN; version 1.1-5) is applied to the matrix of pairwise genomic distances between bases with $p_{n_{i}} \ge t$. The resulting clusters of highly predictive bases identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative base coordinate within each clustered PTBR.

The DBSCAN algorithm has two parameters, *MinPts* and *eps* ($\epsilon$). The *MinPts* parameter was set to the recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. To decide on the optimal value of *t* and $\epsilon$ in *preciseTAD*, we considered the normalized enrichment $N_{e}$ of flanked predicted boundaries, defined as

@@

We present *preciseTAD*, a machine learning approach for precise prediction of TAD or loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on ground truth TAD/loop boundaries obtained from TAD/loop calling tools and genomic annotations as the predictor space. We benchmarked *preciseTAD* against two boundary calling tools, Arrowhead [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We applied *preciseTAD* on cell line-specific boundaries and annotation data from both GM12878 and K562. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and re-sampling techniques. We show that genomic distance between high-resolution boundary regions and transcription factor binding sites, coupled with random under-sampling results in best model performance. We demonstrate that binding of four transcription factors (SMC3, RAD21, CTCF, ZNF143) are sufficient for accurate TAD and loop boundary predictions. Applied to each annotated base, the model predicts probability for each base to be a boundary, and, coupled with spatial clustering (DBSCAN) and scalable partition around medoids clustering (CLARA), allows for the detection of boundaries at the base-level resolution. *preciseTAD* is implemented as an R package, and allows for precise prediction of boundaries for any cell type for which the corresponding genomic annotations are available.

Our machine learning framework yielded several interesting observations. First, we demonstrated that RF models built using *distance*-type predictors outperformed models built on previously published feature engineering techniques including signal strength, overlap counts and overlap percents [@mourad2016computational; @Moore:2015aa; @Al-Bkhetan:2018aa; @kai2018predicting]. We further demonstrated that class imbalance hinders boundary prediction, but can be effectively addressed by a simple random under-sampling technique, an aspect of boundary prediction that was unaddressed in previous studies [@mourad2016computational; @Moore:2015aa; @Al-Bkhetan:2018aa]. We showed that information about only four transcription factors (SMC3, RAD21, CTCF, ZNF143) is necessary and sufficient for accurate boundary prediction, outperforming histone modification- and BroadHMM-built models [@Moore:2015aa; @Al-Bkhetan:2018aa]. These are known components of the loop extrusion model, an established theory for how loops are made by a ring-shaped adenosine triphosphatase complex called cohesin [@hansen2018recent; @sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Interestingly, the same transcription factors accurately predicted both TAD and loop boundaries, suggesting similarity of the mechanisms of TAD and loop formation. This suggested that the random forest model, when tuned and feature engineered correctly, is highly effective in predicting biologically relevant domain boundaries.

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC). Our aim was to have a balanced metric sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs across different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance, preventing us to investigate its effect, and also omits from its calculation TN values. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling. [???refer to supp tables 4 and 5 here when ready]

Guided by both low-resolution Hi-C data and high-resolution genome annotation data, *preciseTAD* predicts base-level resolution boundaries, alleviating resolution limitations of Hi-C data. We show that, unlike Arrowhead, *preciseTAD* does not inflate the number of predicted boundaries, instead, providing only the most biologically meaningful boundaries that demarcate regions of high inter-chromosomal interactions. *preciseTAD* TAD and loop boundaries were enriched for known architectural transcription factors including CTCF, RAD21, SMC3, and ZNF143, supporting recent observations that TADs and loops are dynamic structures generated by similar mechanisms [@hansen2018recent]. Likewise, *preciseTAD* boundaries were found to be more conserved between GM12878 and K562 cell lines, a known feature among the 3D architecture of the human genome, further highlighting their biological significance.

*preciseTAD* offers flexibility in controlling both the number of TAD(loop) boundaries as well as the distance between them. The two primary parameters are the probability threshold (*t*) and $\epsilon$ (refered to as eps; parameter of DBSCAN). The combination of these two quantities change the resulting number of predicted boundaries from *preciseTAD*. A lower threshold and $\epsilon$ will result in more clusters of bases, and therefore, more boundaries. As a heuristic, we evaluated the pairwise combination of 4 different thresholds ($t=(0.95,0.975,0.99,1.0)$) and 4 different $eps$ values ($\epsilon=(r, 2r, 3r, 4r, 5r)$)), where $r=5000$, the optimal resolution. We found that the normalized overlaps--calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries --between top TFBS sites and flanked preciseTAD boundaries were maximized for combinations of t=1.0 and $eps=r$ (Supplementary Figure 9).

A limitation of *preciseTAD* is that it is dependent on "ground truth" boundaries provided by a user-defined domain caller. Given the wide variety of domain callers and their variable performance [@Zufferey:2018aa; @forcato2017comparison], it is hard to define "ground truth" TAD boundaries. <!--Ideally we would benchmark *preciseTAD* against simulated boundaries. While methods for simulating Hi-C data sets with boundary annotations exist [@forcato2017comparison; @zheng2020freehi], methods for simulating the associated genomic annotations (the main component guiding *preciseTAD* predictions) are lacking.-->Ultra-deep Hi-C sequencing [@bonev2017multiscale] and newer technologies for precise mapping of chromatin interactions (e.g., Micro-C [@Krietenstein:2020aa]), coupled with more precise technologies for genomic annotation profiling (e.g., CUT&RUN for precise mapping of transcription factor binding sites) will help to refine the location and the genomic signatures of "ground truth" TAD boundaries. In the current work, given *preciseTAD* can be trained on the whole-genome data, we feel that the total number of TAD/loop boundaries is sufficient to guide learning of the association between genomic annotations and boundaries for precise boundary predictions.

@@

In summary, we demonstrate that TAD/loop boundary prediction is a multi-faceted problem requiring consideration of multiple statistical and biological properties of genomic data. Simply considering the properties of Hi-C contact matrices ignores the fundamental roles of known molecular drivers of the 3D chromatin structures. Instead, we propose *preciseTAD*, a supervised machine learning framework that leverages both Hi-C contact matrix information and genomic annotations. Our method introduces three concepts - *shifted binning*, *distance*-type predictors, and *random undersampling* - which we use to build random forest classification models for predicting boundary regions. Our method is able to bridge the resolution gap between 1D ChIP-seq annotations and 3D Hi-C sequencing data for more precise and biologically meaningful boundary identification. We introduce *preciseTAD*, an open source R package for leveraging random forests to predict TAD/loop boundaries at base resolution. We hope that *preciseTAD* will serve as an efficient and easy-to-use tool to further explore the 3D organization of the genome.

>>>>

To precisely identify boundary locations, we first constructed a _base-level resolution_ predictor space for a particular chromosome *i*. We then apply our optimally trained RF model on the base-level predictor space to extract the probability vector, $p_{n_{i}}$, denoting each base's probability of being a boundary. Here, $n_{i}$ represents the length of chromosome *i*. A threshold ($t$) is specified on $p_{n_{i}}$, such that base coordinates with $p_{n_{i}} \ge t$ are extracted. Next, density-based spatial clustering of applications with noise (DBSCAN; version 1.1-5) is applied to the matrix of pairwise genomic distances between bases with $p_{n_{i}} \ge t$. The resulting clusters of highly predictive bases identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative base coordinate within each clustered PTBR.

The DBSCAN algorithm has two parameters, *MinPts* and *eps* ($\epsilon$). The *MinPts* parameter was set to the recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. To decide on the optimal value of $t$ and $\epsilon$ in *preciseTAD*, we considered the normalized enrichment $N_{e}$ of flanked predicted boundaries, defined as

@@

We present *preciseTAD*, a machine learning approach for precise prediction of TAD or loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on ground truth TAD/loop boundaries obtained from domain calling tools, and genomic annotations as the predictor space. *preciseTAD* was benchmarked against two boundary calling tools, Arrowhead [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and re-sampling techniques. Our results demonstrate that genomic distance between high-resolution boundary regions and transcription factor binding sites, coupled with random under-sampling results in best model performance. We demonstrate that binding of four transcription factors (SMC3, RAD21, CTCF, ZNF143) are sufficient for accurate boundary predictions. Applied to each annotated base, the model predicts probability for each base to be a boundary. Further, coupled with spatial clustering (DBSCAN) and scalable partition around medoids clustering (CLARA), it allows for the detection of boundaries at the base-level resolution. *preciseTAD* is implemented as an R package, and allows for precise prediction of boundaries for any cell type for which the corresponding genomic annotations are available.

Our machine learning framework yielded several interesting observations. First, we demonstrated that RF models built using *distance*-type predictors outperformed models built on previously published feature engineering techniques including signal strength, overlap counts and overlap percents [@mourad2016computational; @Moore:2015aa; @Al-Bkhetan:2018aa; @kai2018predicting]. We further demonstrated that class imbalance hinders boundary prediction, but can be effectively addressed by a simple random under-sampling technique, an aspect of boundary prediction that was unaddressed in previous studies [@mourad2016computational; @Moore:2015aa; @Al-Bkhetan:2018aa]. We showed that information about only four transcription factors (SMC3, RAD21, CTCF, ZNF143) is necessary and sufficient for accurate boundary prediction, outperforming histone modification- and BroadHMM-built models [@Moore:2015aa; @Al-Bkhetan:2018aa]. These are known components of the loop extrusion model, an established theory of how loops are made by a ring-shaped adenosine triphosphatase-driven complex called cohesin [@hansen2018recent; @sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Interestingly, the same transcription factors accurately predicted both TAD and loop boundaries, suggesting similarity of the mechanisms of TAD and loop formation. This suggested that the random forest model, when tuned and feature engineered correctly, is highly effective in predicting biologically relevant domain boundaries.

<!-- Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC). Our aim was to have a balanced metric sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs across different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance, preventing us to investigate its effect, and also omits from its calculation TN values. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling. [???refer to supp tables 4 and 5 here when ready] -->

Guided by both low-resolution Hi-C data and high-resolution genome annotation data, *preciseTAD* predicts base-level resolution boundaries, alleviating resolution limitations of Hi-C data. We show that, unlike Arrowhead, *preciseTAD* does not inflate the number of predicted boundaries, instead, providing only the most biologically meaningful boundaries that demarcate regions of high inter-chromosomal interactions. *preciseTAD* boundaries were enriched for known architectural transcription factors including CTCF, RAD21, SMC3, and ZNF143, supporting recent observations that TADs and loops may be generated by similar mechanisms [@hansen2018recent]. Likewise, *preciseTAD* boundaries were more conserved between GM12878 and K562 cell lines, a known feature among the 3D architecture of the human genome, further highlighting their biological significance.

*preciseTAD* offers flexibility in controlling both the number of predicted boundaries as well as the distance between them. The two primary parameters are the probability threshold $t$ and $\epsilon$ (referred to as eps; parameter of DBSCAN). The combination of these two quantities change the resulting number of predicted boundaries from *preciseTAD*. Lower values of $t$ and $\epsilon$ will result in more clusters of bases, and therefore, more boundaries. As a heuristic, we evaluated the pairwise combination of four different thresholds ($t=(0.95,0.975,0.99,1.0)$) and four different $eps$ values ($\epsilon=(r, 2r, 3r, 4r, 5r)$)), where $r=5000$, the optimal resolution. We found that the normalized overlaps--calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries --between top TFBS sites and flanked preciseTAD boundaries were maximized for combinations of $t=1.0$ and $eps=r$ (Supplementary Figure 9).

A limitation of *preciseTAD* is that it is dependent on "ground truth" boundaries provided by a user-defined domain caller. Given the wide variety of domain callers and their variable performance [@Zufferey:2018aa; @forcato2017comparison], it is hard to define "ground truth" boundaries. <!--Ideally we would benchmark *preciseTAD* against simulated boundaries. While methods for simulating Hi-C data sets with boundary annotations exist [@forcato2017comparison; @zheng2020freehi], methods for simulating the associated genomic annotations (the main component guiding *preciseTAD* predictions) are lacking.-->Ultra-deep Hi-C sequencing [@bonev2017multiscale] and newer technologies for precise mapping of chromatin interactions (e.g., Micro-C [@Krietenstein:2020aa]), coupled with more precise technologies for genomic annotation profiling (e.g., CUT&RUN for precise mapping of transcription factor binding sites) will help to refine the location and the genomic signatures of "ground truth" TAD boundaries. In the current work, given *preciseTAD* can be trained on the whole-genome data, we feel that the total number of TAD/loop boundaries is sufficient to guide learning of the association between genomic annotations and boundaries for precise boundary predictions.

@@

In summary, we demonstrate that TAD/loop boundary prediction is a multi-faceted problem requiring consideration of multiple statistical and biological properties of genomic data. Simply considering the properties of Hi-C contact matrices ignores the fundamental roles of known molecular drivers of the 3D chromatin structures. Instead, we propose *preciseTAD*, a supervised machine learning framework that leverages both Hi-C contact matrix information and genomic annotations. Our method introduces three concepts - *shifted binning*, *distance*-type predictors, and *random undersampling* - which we use to build random forest classification models for predicting boundary regions. Our method is able to bridge the resolution gap between 1D ChIP-seq annotations and 3D Hi-C sequencing data for more precise and biologically meaningful boundary identification. We introduce *preciseTAD*, an open source R package for leveraging random forests to predict TAD/loop boundaries at base resolution. We hope that *preciseTAD* will serve as an efficient and easy-to-use tool to further exploration of the 3D organization of the genome.
